### 2006-01-26 inst/paper/paper2.tex
Added a new version of paper 2.  Must have also made a few small changes to the code.
git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MVCClass@15900 bc3139a8-67e5-0310-9ffc-ced21a209358

\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}

@@

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{December 19, 2005} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in a previous paper, titled ``Creating Linked, Interactive
Views to Explore Multivariate Data'' \cite{EW05}, so the focus in this paper
is on one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis; creating linked,
interactive views of linked data sets.

First brief definitions of linked views, interactive views and linked data
sets are given.  Linked views mean that if a component that represents
data (such as a point on a plot or a row in a spreadsheet) changes its
appearance on one view, then the corresponding component on a second view
also changes its appearance.  For components on different views to be
corresponding, the data displayed in those components must come from the same
entity in a model.  As an example, suppose that the model consisted of four
variables for the fifty states' murder rate, assault rate, rape rate, and
percent of population living in urban areas.  Now the model could be
represented as a matrix of four columns by fifty rows with four columns for
the four variables and fifty rows for the fifty states.  In this model, the
entities are the states.  If two scatterplots were created as views of
this model, where the first plot displayed murder rate versus assault rate and
the second plot displayed rape rate versus percent urban population, then the
points on each plot that referred to the state of Montana are
corresponding components.   Linked views, which have a long history in
statistics, are now considered a standard feature of interactive data
visualization software \cite{GGobi}.  Linked views have been implemented in
several pieces of statistical software, such as XLisp-Stat and GGobi.
Implementing linked views is based on the model-view-controller (MVC) design
\cite{DesignPatterns}, which is a widely used and well understood paradigm.
The MVC design is discussed in more detail in Section \ref{Ssec:Design}.

Interactivity is an idea that is familiar from using the Internet.  For
example, clicking on a hyperlink on a web page results in a new web page
opening.  Another example is clicking on the login button after entering a
user name and password to view one's email.  Interactivity means that there is
a response to some user action.  It is a powerful paradigm and allows users to
quickly navigate and explore complex documents or complex data sets.  As an
example, suppose that a user has two views: one view is a graph of molecular
functions for genes and the other view is a heatmap of gene expression data.
If the user clicks on a node that represents a particular molecular function,
then the response to this action could be to highlit all genes in the heatmap
view that express this molecular function.  This interactivity allows the
user to quickly see if genes with the same molecular function have similar
expression levels.

One particular model for creating interactive applications involves three
components: some user action or input, which is referred to as an event;
a response to that action, which is executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which is referred to as the signal handler.  The events can
range from the user moving the cursor over a plot, to pressing a button on a
graphical user interface (GUI), to pressing certain keys.  Events are
interpreted based on which window has focus.  For example, in the XWindows
system, if a user types C on the keyboard and the window that has focus is a
text editor, then the letter C is printed on the text editor's screen.
Any other open windows that do not have
focus do not respond to this event.  The callback function is a unique
function because its signature is determined by the event that the callback
function is linked to.  The parameters of the callback function may include
information about the event, such as the on screen location of the event, and
information about the object that had focus when the event occurred, such as a
button on the GUI.  The signal handler is the piece that must look to see if
an event has occurred and then when the event has occurred, the signal handler
calls the appropriate callback function.

@@

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.6}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

@@

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

In the design presented in this paper, the components consist of MVC objects
and communication within and between MVC objects is performed by message
objects.  This solution uses the power and simplicity of the MVC design and
expands it to the multiple data set problem.  Now we can build on the original
MVC design, which has been well tested, instead of starting a new design from
scratch.  The strength of this multiple linked MVC design is that it uses MVC
components, which can be individually created and then linked.  Thus, these
components are reusable.  These design decisions are discussed further in
Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked Views of Linked Data Sets. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the MVC paradigm is discussed.  However, the limitation with this design is
that it expects one data set, which limits the type of data we can visualize.
There are many situations where multiple data sets occur, particularly any
time there are experimental or study data linked to meta-data.  Being able to
visualize data from multiple data sets is helpful to understand the
relationships between variables.  Thus, after discussing the limitations of
the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
linked MVC objects.  The design of individual MVC objects is discussed in
Section \ref{Sec:OneMVC} while the implementation of linked MVC objects is
discussed in Section \ref{Sec:MultMVC}.  The goals of the software discussed
in this paper are to create linked views of linked data sets, to create
interactive views where the response to an event can be changed, and to create
an extensible design that allows future users to make additions.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.6}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

@@

A more straightforward approach may be to link MVCs.  Each MVC is a coherent
object and the interactions between MVCs are encapsulated in a set of
messages.  We use one MVC object to handle each data set.  Thus, within a MVC
object, one can create linked views of that particular data set, using the
standard MVC design.  But between MVC objects we design a second messaging
system so that a change in one data set is sent to the other linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the StateCity table that links cities
with the states they reside in.  Now for these two data sets there are
two MVC objects.  One MVC object for the State data and its views and one MVC
object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two MVC Objects. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

Figure \ref{Fig:MultMVC} shows one view for the State data and one view
for the City data.  You can see how the State and City data are linked based
on the coloring of the points in the plots.  The points referring to the state
of Pennsylvania are red, the points referring to California are blue, and the
points referring to Minnesota are green.  Even though there are two MVC
objects here, with one MVC object for each data set, it is still clear that
the views are linked through the data sets and that is shown in this example
through the coloring.

Not only does this solution use the power and simplicity of the MVC design, but
this multiple MVC design is structured to fit the linking seen in the multiple
data sets.  Clearly, the design should be built around the problem and
multiple MVC objects with messaging between the MVC objects fits the problem
of creating linked views of multiple linked data sets.  In designing an
appropriate messaging system between MVC objects, consideration is given to
what information should pass between MVC objects and to how messages propagate
between MVC objects.  As for the relationship between the MVC objects, it
should reflect the relationship between the data sets.

Although two examples of linked data sets have been discussed thus far,
this paper focuses on experimental data that are linked to meta-data, as
the data we are most interested in visualizing.  In this situation, the
experimental data can be thought of as determining what meta-data are studied.
In the example of microarray experiment data linked to Gene Ontology
meta-data, only Gene Ontology data that refer to genes in the microarray
experiment are included in the meta-data.  Thus, there is a hierarchical
relationship between these data sets and the MVC objects that contain them,
where one MVC object creates another MVC object.  The original MVC
object is the parent MVC and the new MVC object that is created from the
parent MVC is the child MVC.  These decisions are discussed in the following
sections as the design for multiple MVC objects is explained.

@@

Implementing linked, interactive views of linked data sets is done using an
object-oriented programming design for extensibility.  The \Rpackage{MVCClass}
package implements the design of the MVC classes and generic functions.  The
\Rpackage{MVCClass} package defines the individual model, view, and
controller classes as well as the \Robject{MVC} class that binds the
individual model, view, and controller classes together into one unit.  The
package also defines the message classes that allow the components to
communicate with each other.  Thus, the \Rpackage{MVCClass} package
contains the classes necessary for creating linked views.   The goal of
placing these class and generic function definitions into one package is
that they are available for other R packages that implement linked views
of data, such as the \Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.

@@

The next subsections explain the class definitions and the methods that
pertain to a single MVC object.  Thus, the model classes,
the view classes, the controller, the \Robject{MVC} class (which binds the
model, view and controller classes together), and the message classes that
pass information between the three components in one MVC object are
discussed in the following subsections.  All of the class definitions are
implemented in the \Rpackage{MVCClass} package and all of the methods are
defined in the \Rpackage{iSNetwork} package.  Message class definitions that
pertain to communication between the MVC objects are discussed in Section
\ref{Sec:MultMVC}.

@@

Figure \ref{Fig:Model} shows the inheritance structure for the model classes.
Here the virtual class, \Robject{gModel}, has the slots \Rslot{modelData},
\Rslot{linkData}, \Rslot{virtualData}, and \Rslot{modelName}.  These four
slots are common information that all models need to store.  The
\Rslot{modelData} slot is the data set for this model, the \Rslot{linkData}
slot is a list of two functions, \Rfunction{toParent} and
\Rfunction{fromParent}, which let this model send information to its parent
model and receive information from its parent model, respectively (see
Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
parent and child models are linked), the \Rslot{virtualData} slot is data
pertaining to the views that needs to be stored with the model so that it can
be shown in all views of this model, and the \Rslot{modelName} slot is the
name of the model.

@@

We encapsulate the particular data model in the class definition, which allows
for efficient dispatch, and accessor functions that are appropriate for the
underlying implementation.  The specific model classes used in the
\Rpackage{iSNetwork} package to represent data are \Robject{dfModel},
\Robject{graphModel}, and \Robject{exprModel}.  The \Robject{dfModel} class
represents a model where the \Rslot{modelData} slot has class data frame (or
matrix), the \Robject{graphModel} class represents a model where the
\Rslot{modelData} slot has class graph, and the \Robject{exprModel} class
represents a model where the \Rslot{modelData} slot has class exprSet, a class
used to represent gene expression data \cite{BioC}.

@@

Although only three model classes are currently implemented, future users are
welcome to make additions to the model classes for new data structures they
want to represent.  For example, a user may want to add a new model
class called \Robject{tsModel} to represent time series data.  To make this
addition, the user extends the \Robject{gModel} class in his or her own
package.

@@

The view classes represent the visual depictions of the model.  All views
need to store some common information and respond to certain events through
methods.  This consideration leads to an object model where the different view
classes inherit from a view virtual class, called \Robject{genView}.  The
object model for the view classes is shown in Figure \ref{Fig:View}.

@@

The \Robject{genView} class has three slots, \Rslot{dataName}, \Rslot{win},
and \Rslot{winNum}.  \Rslot{dataName} is the name of the model that the view
displays, \Rslot{win} is the Gtk window object that holds the view, and
\Rslot{winNum} is the number of the window so that the window can be
identified in the Window menu on the GUI (see Section \ref{Ssec:OneCont}).
All views need to know these three pieces of information so
\Robject{genView} binds the view classes together.

As for specific view classes, the \Robject{spreadView} class represents a
spreadsheet view of the data.  This view only makes sense for models that
are a two dimensional data structure, such as a matrix or a data frame.  The
information needed for this view is stored in the slot, \Rslot{clist}, which
is the Gtk spreadsheet object.

When creating a plot of a data set, there is a general class, called
\Robject{plotView}, that has the slots, \Rslot{plotDevice}, \Rslot{plotPar},
and \Rslot{drArea}, which store the device number of the plot, the plotting
parameters, and the Gtk drawing area object, respectively.  This class is a
virtual class as it represents a general plot.  The instantiable plot classes
are the specific plot classes, \Robject{sPlotView}, \Robject{graphView}, and
\Robject{heatmapView}, which inherit from the \Robject{plotView} class.

The \Robject{sPlotView} class represents a scatterplot view and it has the
extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
\Rslot{dfRows} are the row names from the model that are shown in the plot,
\Rslot{xvar} is the variable name from the model that is shown as the x
variable in the plot, and \Rslot{yvar} is the variable name from the model
that is shown as the y variable in the plot.  The \Robject{graphView} class
represents a graph plot view and it contains no extra slots besides those in
the \Robject{plotView} class.  The \Robject{heatmapView} represents a heatmap
view of the data and it contains the slot, \Rslot{ordering}, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the ordering in the row and column dendrograms.  Note that some of these
views are only applicable for certain model types.  For example, the
\Robject{graphView} class only makes sense for a view of a graph object,
which is stored in the \Robject{graphModel} class.

@@

For \Robject{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a mouse button has been pressed and the
\Rfunction{motionEvent} method is called when Gtk emits a signal that
the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods for a \Robject{sPlotView} object
determine which point on the plot was clicked or under the cursor,
respectively, using the data in the \Rslot{where} parameter.  For
\Robject{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a row in the spreadsheet has been selected or unselected
and the \Rfunction{clickEvent} method determines which row was changed.
For \Robject{graphView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a mouse button has been pressed and the
\Rfunction{motionEvent} is called when Gtk emits a signal that the cursor has
been moved over the view.  Here, though, the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods determine which node was clicked or under
the cursor, respectively, using the data in the \Rslot{where} parameter.

The \Robject{heatmapView} class is different from the other views in that it
has neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method.  The
current implementation of the \Rfunction{heatmap} function does not return
information about where plot objects are placed on the device.  As a result,
it is not possible to convert the user coordinates of where the event occurred
into an object (such as the row and column coordinates of the heatmap) on
the view.  Thus, the \Robject{heatmapView} class is the only view that is not
currently interactive.

@@

The inheritance structure of the view classes is intended to be extensible.
For instance, if users wanted to add a new plot view, say a density plot, they
could create a new view class that inherits from \Robject{plotView}, called
\Robject{dPlotView}.  This new class, \Robject{dPlotView}, has slots
named \Rslot{dfRows} and \Rslot{col} to store which rows and which column,
respectively, were used to calculate the density shown in the view.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{ControlWindow.jpg}}
    \caption{ The Control Window. }
    \label{Fig:ContWin}
  \end{center}
\end{figure}

>>>>

\newcommand{\Rfunction}[1]{{\textbf{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rslot}[1]{\textsl{#1}}
\newcommand{\Rclass}[1]{\textsf{#1}}

@@

\begin{titlepage}
\begin{center}
\Large
\vspace*{1in}
\textbf{Creating Linked, Interactive Views to \linebreak Explore Linked Data} \\
\vspace{1in}
\textbf{Dana-Farber Cancer Institute}\\ 
\textbf{January 25, 2006} \\
\vspace{1in} 
\textbf{Elizabeth Whalen} \\
\end{center}
\end{titlepage}
\normalsize
\newpage

@@

There are many ways to visualize multivariate data where more than
two or three dimensions are needed to view the data.  Several of these ideas
were discussed in previous work \cite{EW05}, so the focus in this paper
is on one specific method of visualizing multivariate data that is
particularly useful for exploratory data analysis; creating linked,
interactive views of linked data sets.

First definitions of linked views, interactive views and linked data
sets are given in the following paragraphs.  Linked views mean that if a
component that represents data (such as a point on a plot or a row in a
spreadsheet) changes its appearance on one view, then the corresponding
component on a second view also changes its appearance.  For components on
different views to be corresponding, the data displayed in those components
must come from the same entity in a model.  As an example, suppose that the
model consists of the gene expression levels from four samples.  If the same
one hundred genes are studied in the four samples, then the model could be
represented as a matrix of four columns by one hundred rows with four columns
for the four samples and one hundred rows for the one hundred genes.  In this
model, the entities are the genes.  If two scatterplots are created as views
of this model, where the first plot displays sample one versus sample two and
the second plot displays sample three versus sample four, then the points on
each plot that referred to the same gene are corresponding components.  Linked
views, which have a long history in statistics, are now considered a standard
feature of interactive data visualization software \cite{GGobi}.  Linked views
have been implemented in several pieces of statistical software, such as
XLisp-Stat \cite{Lisp} and GGobi \cite{GGobiMan}.  Implementing linked views
is based on the model-view-controller (MVC) design \cite{DesignPatterns},
which is a widely used and well understood paradigm.  The MVC design is
discussed in more detail in Section \ref{Ssec:Design}.

Interactivity is an idea that is familiar from using an Internet browser.
Clicking on a hyperlink on a web page results in a new web page
opening.  Another example is clicking on the login button after entering a
user name and password to view one's email.  Interactivity means that there is
a response to some user action.  It is a powerful paradigm and allows users to
quickly navigate and explore complex documents or complex data sets.  As an
example, suppose that a user has two views: one view is a graph representing
the molecular function of genes and the other view is a heatmap of gene
expression data.  If the user clicks on a node that represents a particular
molecular function, then the response to this action could be to highlight all
genes in the heatmap view that correspond to this molecular function.  This
interactivity allows the user to quickly see if genes with the same molecular
function have similar expression levels.

One particular model for creating interactive applications involves three
components: some user action or input, which is referred to as an event;
a response to that action, which is executed by a function that is called
a callback function; and a method that connects the event to the callback
function, which is referred to as the signal handler.  The events can
range from the user moving the cursor over a plot, to pressing a button on a
graphical user interface (GUI), to pressing certain keys.  Events are
interpreted based on which window has focus.  For example, in the XWindows
system, if a user types C on the keyboard and the window that has focus is a
text editor, then the letter C is printed on the text editor's screen.
Any other open windows that do not have focus do not respond to this event.
The callback function is a unique function because its signature is determined
by the event that the callback function is linked to.  The parameters passed
to the callback function may include information about the event, such as the
on screen location of the event, and information about the object that had
focus when the event occurred, such as a button on the GUI.  The signal
handler is the piece that must look to see if an event has occurred and then
when the event has occurred, the signal handler calls the appropriate callback
function.

@@

\begin{figure}[!h]
  \begin{center}    
    \scalebox{0.55}{\includegraphics{Event.jpg}}
    \caption{ Flow Chart of the Response to an Event. }
    \label{Fig:Event}
  \end{center}
\end{figure}

@@

As a motivating example of linked views based on linked data sets, consider
Figure \ref{Fig:firstMP}.  In this figure, there are two views, one view of
data set 1 and one view of data set 2.  Data set 1 and data set 2 are linked.
When a user initially interacts with the scatterplot view of data set 2, the
response to this interaction is to tell data set 2 to update itself (this is
indicated on the figure as a circled 1).  Then in Step 2, data set 2 tells the
scatterplot view that is should update itself in response to the change in the
data.  Next, in Step 3, because data set 2 has changed and because it is
linked to data set 1, data set 2 sends information to data set 1 about the
change.  In Step 4, the information from data set 2 is converted into
information that data set 1 can understand.  In Step 5, this information is
received by data set 1 and data set 1 updates its data.  Finally, in Step 6,
data set 1 tells its view to update itself.  This example shows the components
and the communication between the components that are necessary to create
linked views of underlying linked data sets.

In the design presented in this paper, the components consist of \Robject{MVC}
objects and communication within and between \Robject{MVC} objects is
performed by message objects.  This solution uses the power and simplicity of
the MVC design and expands it to the multiple data set problem.  Now we can
build on the original MVC design, which has been well tested, instead of
starting a new design from scratch.  The strength of this multiple linked MVC
design is that it uses MVC components, which can be individually created and
then linked.  Thus, these components are reusable.  These design decisions are
discussed further in Sections \ref{Sec:OneMVC} and \ref{Sec:MultMVC}.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.55}{\includegraphics{firstMessagePassing.jpg}}
    \caption{ Linked views of linked Data Sets.  When a user interacts with
      the y versus x scatterplot, the first step, indicated by a circled one,
      is to tell data set 2 to update.  Then all views of data set 2 must be
      updated.  Next any data that are linked to data set 2 must be notified
      that data set 2 changed.  This information from data set 2 must be
      altered so that it can be understood by linked data sets, which is shown
      in the fourth step.  Finally, data set 1 is notified that it should
      update in response to the change in data set 2 and then all views of
      data set 1 are updated. }
    \label{Fig:firstMP}
  \end{center}
\end{figure}

In Section \ref{Ssec:Design}, the design of linked, interactive views based on
the MVC paradigm is discussed.  However, the limitation with this design is
that it expects one data set, which limits the type of data we can visualize.
There are many situations where multiple data sets occur, particularly any
time there are experimental or study data linked to meta-data.  Being able to
visualize data from multiple data sets is helpful to understand the
relationships between variables.  Thus, after discussing the limitations of
the MVC design in Section \ref{Ssec:Limit}, an extension of the MVC design is
discussed in Section \ref{Sec:Extend}.  This extension supports the creation of
linked \Robject{MVC} objects.  The design of individual \Robject{MVC} objects
is discussed in 
Section \ref{Sec:OneMVC} while the implementation of linked \Robject{MVC}
objects is discussed in Section \ref{Sec:MultMVC}.  The goals of the software
discussed in this paper are to create linked views of linked data sets, to
create interactive views where the response to an event can be changed, and to
create an extensible design that allows future users to make additions.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.5}{\includegraphics{ExMVC.jpg}}
    \caption{ A Model with its Views. }
    \label{Fig:ExMVC}
  \end{center}
\end{figure}

@@

A more straightforward approach may be to link \Robject{MVCs}.  Each
\Robject{MVC} is a coherent object and the interactions between \Robject{MVCs}
are encapsulated in a set of messages.  We use one \Robject{MVC} object to
handle each data set.  Thus, within a \Robject{MVC} object, one can create
linked views of that particular data set, using the standard MVC design.  But
between \Robject{MVC} objects we design a second messaging system so that a
change in one data set is sent to the other linked data sets.

@@

\normalsize
As an example, suppose that one has two data sets: one with information at the
state level and one with information at the city level.  These tables are
shown in Table \ref{Tab:CityState}.  The State data include Life Expectancy
(in years from 1970) and Income (per capita from 1974) data
while the City data include Precipitation (in inches from 1975) and Average
High Temperature in January (in Fahrenheit from 2005)
data.  These two data sets are linked by the StateCity table that links cities
with the states they reside in.  Now for these two data sets there are
two \Robject{MVC} objects.  One \Robject{MVC} object for the State data and
its views and one \Robject{MVC} object for the City data and its views.

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.7}{\includegraphics{MultipleMVC.jpg}}
    \caption{ Example of Two \Robject{MVC} Objects.  These two \Robject{MVC}
      objects are linked by the StateCity table and the color of the points
    indicates the linking between the data.  The points referring to the state
    of Pennsylvania are red, the points referring to the state of California
    are blue, and the points referring to Minnesota are green. }
    \label{Fig:MultMVC}
  \end{center}
\end{figure}

Figure \ref{Fig:MultMVC} shows one view for the State data and one view
for the City data.  You can see how the State and City data are linked based
on the coloring of the points in the plots.  The points referring to the state
of Pennsylvania are red, the points referring to California are blue, and the
points referring to Minnesota are green.  Even though there are two
\Robject{MVC} objects here, with one \Robject{MVC} object for each data set,
it is still clear that the views are linked through the data sets and that is
shown in this example through the coloring.

Not only does this solution use the power and simplicity of the MVC design, but
this multiple MVC design is structured to fit the linking seen in the multiple
data sets.  Clearly, the design should be built around the problem and
multiple \Robject{MVC} objects with messaging between the \Robject{MVC}
objects fits the problem of creating linked views of multiple linked data
sets.  In designing an appropriate messaging system between \Robject{MVC}
objects, consideration is given to what information should pass between
\Robject{MVC} objects and to how messages propagate between \Robject{MVC}
objects.  As for the relationship between the \Robject{MVC} objects, it should
reflect the relationship between the data sets.

Although two examples of linked data sets have been discussed thus far,
this paper focuses on experimental data that are linked to meta-data, as
the data we are most interested in visualizing.  In this situation, the
experimental data can be thought of as determining what meta-data are studied.
In the example of microarray experiment data linked to Gene Ontology
meta-data, only Gene Ontology data that refer to genes in the microarray
experiment are included in the meta-data.  Thus, there is a hierarchical
relationship between these data sets and the \Robject{MVC} objects that
contain them, where one \Robject{MVC} object creates another \Robject{MVC}
object.  The original \Robject{MVC} object is the parent \Robject{MVC} and the
new \Robject{MVC} object that is created from the parent \Robject{MVC} is the
child \Robject{MVC}.  These decisions are discussed in the following sections
as the design for multiple \Robject{MVC} objects is explained.

@@

Implementing linked, interactive views of linked data sets is done using an
object-oriented programming design for extensibility.  The \Rpackage{MVCClass}
package implements the design of the MVC classes and generic functions.  The
\Rpackage{MVCClass} package defines the individual model,
view, and controller classes as well as the \Rclass{MVC}
class that binds the individual model, view, and
controller classes together into one unit.  The package also defines
the message classes that allow the components to communicate with
each other.  Thus, the \Rpackage{MVCClass} package contains the classes
necessary for creating linked views.   The goal of placing these class and
generic function definitions into one package is that they are available for
other R packages that implement linked views of data, such as the
\Rpackage{iSPlot} and the \Rpackage{iSNetwork} packages.

@@

The next subsections explain the class definitions and the methods that
pertain to a single \Robject{MVC} object.  Thus, the model classes,
the view classes, the controller, the \Rclass{MVC} class
(which binds the model, view and controller classes
together), and the message classes that pass information between the
three components in one \Robject{MVC} object are discussed in the following
subsections.  All of the class definitions are implemented in the
\Rpackage{MVCClass} package and all of the methods are defined in the
\Rpackage{iSNetwork} package.  Message class definitions that pertain
to communication between the \Robject{MVC} objects are discussed in Section
\ref{Sec:MultMVC}.

@@

Figure \ref{Fig:Model} shows the inheritance structure for the model
classes. Here the virtual class, \Rclass{gModel}, has the slots
\Rslot{modelData}, \Rslot{linkData}, \Rslot{virtualData}, and
\Rslot{modelName}.  These four slots are common information that all models
need to store.  The \Rslot{modelData} slot is the data set for this model, the
\Rslot{linkData} slot is a list of two functions, \Rfunction{toParent} and
\Rfunction{fromParent}, which let this model send information to its parent
model and receive information from its parent model, respectively (see
Sections \ref{Ssec:OneMVC} and \ref{Ssec:MultLink} for more information on how
parent and child models are linked), the \Rslot{virtualData} slot is data
pertaining to the views that needs to be stored with the model so that it can
be shown in all views of this model, and the \Rslot{modelName} slot is the
name of the model.

@@

We encapsulate the particular data model in the class definition, which allows
for efficient dispatch, and accessor functions that are appropriate for the
underlying implementation.  The specific model classes used in the
\Rpackage{iSNetwork} package to represent data are \Rclass{dfModel},
\Rclass{graphModel}, and \Rclass{exprModel}.  The \Rclass{dfModel} class
represents a model where the \Rslot{modelData} slot has class data frame (or
matrix), the \Rclass{graphModel} class represents a model where the
\Rslot{modelData} slot has class graph, and the \Rclass{exprModel} class
represents a model where the \Rslot{modelData} slot has class exprSet, a class
used to represent gene expression data \cite{BioC}.

@@

Although only three model classes are currently implemented, future
users are welcome to make additions to the model classes for new data
structures they want to represent.  For example, a user may want to add a new
model class called \Rclass{tsModel} to represent time series data.
To make this addition, the user extends the \Rclass{gModel} class in his or
her own package.

@@

The view classes represent the visual depictions of the model.  All
views need to store some common information and respond to certain events
through methods.  This consideration leads to an object model where the
different view classes inherit from a view virtual class,
called \Rclass{genView}.  The object model for the view classes is
shown in Figure \ref{Fig:View}.

@@

The \Rclass{genView} class has three slots, \Rslot{dataName}, \Rslot{win},
and \Rslot{winNum}.  \Rslot{dataName} is the name of the model that the view
displays, \Rslot{win} is the Gtk window object that holds the view, and
\Rslot{winNum} is the number of the window so that the window can be
identified in the Window menu on the GUI (see Section \ref{Ssec:OneCont}).
All views need to know these three pieces of information so
\Rclass{genView} binds the view classes together.

As for specific view classes, the \Rclass{spreadView} class
represents a spreadsheet view of the data.  This view only makes sense for
models that are a two dimensional data structure, such as a matrix or a data
frame.  The information needed for this view is stored in the slot,
\Rslot{clist}, which is the Gtk spreadsheet object.

When creating a plot of a data set, there is a general class, called
\Rclass{plotView}, that has the slots, \Rslot{plotDevice}, \Rslot{plotPar},
and \Rslot{drArea}, which store the device number of the plot, the plotting
parameters, and the Gtk drawing area object, respectively.  This class is a
virtual class as it represents a general plot.  The instantiable plot classes
are the specific plot classes, \Rclass{sPlotView}, \Rclass{graphView}, and
\Rclass{heatmapView}, which inherit from the \Rclass{plotView} class.

The \Rclass{sPlotView} class represents a scatterplot view and it has the
extra slots, \Rslot{dfRows}, \Rslot{xvar}, and \Rslot{yvar}, where
\Rslot{dfRows} are the row names from the model that are shown in the plot,
\Rslot{xvar} is the variable name from the model that is shown as the x
variable in the plot, and \Rslot{yvar} is the variable name from the model
that is shown as the y variable in the plot.  The \Rclass{graphView} class
represents a graph plot view and it contains no extra slots besides those in
the \Rclass{plotView} class.  The \Rclass{heatmapView} represents a heatmap
view of the data and it contains the slot, \Rslot{ordering}, which is a list
that is returned from the \Rfunction{heatmap} function to give information
about the ordering in the row and column dendrograms.  Note that some of these
views are only applicable for certain model types.  For example, the
\Rclass{graphView} class only makes sense for a view of a graph object,
which is stored in the \Rclass{graphModel} class.

@@

For \Rclass{sPlotView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a mouse button has been pressed and the
\Rfunction{motionEvent} method is called when Gtk emits a signal that
the cursor has been moved over the view.  Both the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods for a \Robject{sPlotView} object
determine which point on the plot was clicked or under the cursor,
respectively, using the data in the \Rslot{where} parameter.  For
\Rclass{spreadView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a row in the spreadsheet has been selected or unselected
and the \Rfunction{clickEvent} method determines which row was changed.
For \Rclass{graphView}, the \Rfunction{clickEvent} method is called when Gtk
emits a signal that a mouse button has been pressed and the
\Rfunction{motionEvent} is called when Gtk emits a signal that the cursor has
been moved over the view.  Here, though, the \Rfunction{clickEvent} and
\Rfunction{motionEvent} methods determine which node was clicked or under
the cursor, respectively, using the data in the \Rslot{where} parameter.

The \Rclass{heatmapView} class is different from the other views in that it
has neither a \Rfunction{clickEvent} or a \Rfunction{motionEvent} method.  The
current implementation of the \Rfunction{heatmap} function does not return
information about where plot objects are placed on the device.  As a result,
it is not possible to convert the user coordinates of where the event occurred
into an object (such as the row and column coordinates of the heatmap) on
the view.  Thus, the \Rclass{heatmapView} class is the only view that is not
currently interactive.

@@

The inheritance structure of the view classes is intended to be
extensible.  For instance, if users wanted to add a new plot view, say a
density plot, they could create a new view class that inherits from
\Rclass{plotView}, called \Rclass{dPlotView}.  This new class,
\Rclass{dPlotView}, has slots named \Rslot{dfRows} and \Rslot{col} to store
which rows and which column, respectively, were used to calculate the density
shown in the view.

@@

\begin{figure}[ht]
  \begin{center}
    \scalebox{0.35}{\includegraphics{ControlWindow.jpg}}
    \caption{ The Control Window. }
    \label{Fig:ContWin}
  \end{center}
\end{figure}
